Social Attention for Autonomous Decision-Making in Dense Traffic

### 2019-11-26 paper/main.tex
Updates from Overleaf

% ready for submission
\usepackage[nonatbib]{neurips_2019_ml4ad}

@@

% to compile a camera-ready version, add the [final] option, e.g.:
% \usepackage[final]{neurips_2019_ml4ad}

@@

\begin{abstract}
  We study the design of learning architectures for behavioural planning in a dense traffic setting. Such architectures should deal with a varying number of nearby vehicles, be invariant to the ordering chosen to describe them, while staying accurate and compact. We observe that the two most popular representations in the literature do not fit these criteria, and perform badly on an complex negotiation task. We propose an attention-based architecture that satisfies all these properties and explicitly accounts for the existing interactions between the traffic participants. We show that this architecture enjoys significant performance gains, and that it is able to capture interactions patterns that can be visualized and qualitatively interpreted. Videos and code are available at \url{https://social-attention.github.io/}
\end{abstract}

@@

A popular way to address this limitations is to use a \emph{spatial grid} representation. Instead of explicitly representing spatial information as variables $x, y$ along with other features $f$ directly inside a state $\{s_i=(x_i,y_i,f_i)\}_{i\in[0,N]}$ indexed on the vehicles, they are instead represented implicitly through the layout of several feature variables $f_{ij}$ organized in a tensor structure, where the $(i,j)$ indexes refer to a quantization of the 2D-space. This representation is illustrated in \autoref{fig:representation} (right). Note that the size of this tensor is related to the area covered divided by the quantization step, which reflects a trade-off between accuracy and dimensionality.
In an occupancy grid, the $f$ features contains presence information (0-1) and additional channels such as velocity and heading, as in \citep[e.g.][]{Isele2018, Fridman2018, Bansal2018, Rehder2017c}. Another example is the use of top-view RGB images \citep[e.g.][]{Bagnell2010, Rehder2017, Rehder2017c, Liu2018}.

@@

The proposed architecture is presented in \autoref{fig:architecture}. It is used to represent the $Q$-function that will be optimized by the DQN algorithm. It is composed of a first linear encoding layer whose weights are shared between all vehicles. At that point, the embeddings only contain individual features of size $d_x$. They are then fed to an ego-attention layer, composed of several heads stacked together. The \emph{ego} prefix highlights that it similar to a multi-head self-attention layer \citep{Vaswani2017} but with only a single output corresponding to the ego-vehicle. Such an ego-attention head is illustrated in \autoref{fig:ego-attention} and works in the following way: in order to select a subset of vehicles depending on the context, the ego-vehicle  first emits a single query $Q = [q_0]\in\Real^{1 \times d_k}$, computed with a linear projection $L_q\in\Real^{d_x \times d_k}$ of its embedding. This query is then compared to a set of keys $K = [k_0, \dots, k_N]\in\Real^{N \times d_k}$ containing descriptive features $k_i$ for each vehicle, again computed with a shared linear projection $L_k\in\Real^{d_x \times d_k}$. The similarity between the query $q_0$ and any key $k_i$ is assessed by their dot product $q_0 k_i^T$. These similarities are then scaled by the inverse-square-root-dimension $1/\sqrt{d_k}$\footnote{This scaling is due to the fact that the dot-product of two independent random vectors with mean 0,  variance 1, and dimension $d_k$, is a random variable with mean 0 and variance $d_k$} and normalized with a softmax function $\sigma$ across vehicles. We obtain a stochastic matrix called the \emph{attention matrix}, which is finally used to gather a set of output value $V = [v_0, \dots, v_N]$, where each value $v_i$ is a feature computed with a shared linear projection $L_v\in\Real^{d_x \times d_v}$. Overall, the attention computation for each head can be written as:
\begin{equation}
\text{output}=\underbrace{\sigma\left(\frac{QK^T}{\sqrt{d_k}}\right)}_{\text{attention matrix}}V
\label{eq:selfattention}
\end{equation}
The outputs from all heads are finally combined with a linear layer, and the resulting tensor is then added to the ego encoding as in residual networks. We can easily see that this process is permutation invariant: indeed, a permutation $\tau$ will change the order of the rows in keys $K$ and values $V$ in \eqref{eq:selfattention} but will keep their correspondence. The final result is a dot product of values and key-similarities, which is independent of the ordering.

@@

In this application, we use the \href{https://github.com/eleurent/highway-env}{highway-env} environment \citep{highway-env} for simulated highway driving and behavioural decision-making. We propose a new task where vehicle-to-vehicle interaction plays a significant part: crossing a four-way intersection.
The scene -- composed of two roads crossing perpendicularly -- is populated with several traffic participants initialized with random positions, velocities, and destinations. As in \citep{highway-env}, these vehicle are simulated with the Kinematic Bicycle Model, their lateral control is achieved by a low-level steering controller tracking a target route, and their longitudinal behaviour follows the Intelligent Driver Model \citep{Treiber2000}. However, this model only considers same-lane interactions and special care was required to prevent lateral collisions at the intersection. To that end, we implemented the following simplistic behaviour: each vehicle predicts the future positions of its neighbours over a three-seconds horizon by using a constant velocity model. In case of predicted collision with a neighbour, the yielding vehicle is determined based on road priorities and brakes until the collision prediction ceases.

@@

We evaluate three different agents, whose characteristics are summarized in \autoref{tab:agents}.
\begin{itemize}
	\item \MLPL: a \emph{list of features} state representation is used, as described in \autoref{sec:background}. The model is a simple multi-layer perceptron (MLP). Because this architecture requires a fixed-size input, we use zero-padding to fill the input tensor up to a maximum number $N=14$ of observed vehicles, and add an additional \emph{presence} feature to the coordinates described in \eqref{eq:coordinates} so as to identify active rows.
	\item \MLPG: a \emph{spatial grid} representation is used, as described in \autoref{sec:background}, with a $11 \times 11$ grid where each cell represents a $5\text{m}\times 5$m square. The model is a multi-layer perceptron.
	\item \EgoAtt: a \emph{list of features} state representation is used along with the Ego-Attention architecture described in \autoref{sec:architecture}. As this model supports varying-size inputs, zero-padding is not required.
\end{itemize}

\begin{table}[tp]
	\centering
	\begin{threeparttable}
		\caption{Characteristics of the agents}
		\label{tab:agents}
		\begin{tabular}{lccc}
			\toprule
			Architecture & \MLPL & \MLPG & \EgoAtt \\
			\midrule 
			Input sizes & [15, 7] & [11, 11, 5] & [1-15, 7] \\
			Layers sizes & [128, 128] &  [48, 48] & \makecell[tl]{Encoder: [64, 64] \\Attention: 2 heads, $d_k=32$ \\ Decoder: [64, 64]} \\
			Number of parameters & 3.0e4 & 3.2e4 & 3.4e4 \\
			Variable input size & No & No &  {Yes}  \\
			Permutation invariant & No & {Yes} &  {Yes} \\
			\bottomrule
		\end{tabular}
	\end{threeparttable}
\end{table}

@@

We plot in \autoref{fig:results} the evolution of the total reward, episode length and average velocity during training, over 4000 episodes and repeated across 120 random seeds.
The \MLPL agent learns to accelerate to earn short-term rewards, as shown by its high average velocity, but fails to exploit the information of other vehicles and crashes often, leading to short episodes. We obtain a risky and blind policy that is the worst performing.
Conversely, the \MLPG architecture benefits from its invariance to permutations and manages to learn to brake upon arrival at the intersection to avoid collisions, as we can see from its higher episode length. However, it only proceeds when the intersection has been fully cleared, as reflected by its low average velocity. This results in an overly cautious policy -- a common trait colloquially known as the \emph{freezing robot problem} \citep{Trautman2010} -- with a slight increase in performance.
In stark contrast, the \EgoAtt policy quickly learns both when it must slow down at the intersection (see the high episode length), but also when it can exploit the gaps in the traffic and take way to vehicles that are far or slow enough (see the higher average velocity than \MLPG). This translates as a significant performance improvement, and the overall resulting behaviour is qualitatively more nuanced and human-like.

@@

In any given state, the attention matrix can be visualized in the following way: we connect the ego-vehicle to every vehicle by a line of width proportional to the corresponding attention weight. Since the architecture can contain several ego-attention heads, we use different colours to distinguish them. In our experiments, two attention heads were used and will be represented in green and blue. We observe in \autoref{fig:heads} that they specialized to focus on different areas: the green head is only watching the vehicles coming from the left, while the blue head restricts itself to vehicles in the front and right directions. However, we notice that both heads exhibit a common behaviour: they direct their attention to incoming vehicles that are likely to collide with the ego-vehicle, depending on their current position, heading, velocity, and ignore those that are too far or in a conflict-less situation. In particular, the attention tends to increase when vehicles get closer, as shown in \autoref{fig:distance}. It can also be very sensitive to small variations in the traffic state, as reflected in \autoref{fig:sensitivity}. A full episode showcasing interactions with several vehicles is shown in \autoref{fig:episode}.

\begin{figure}[tp]
	\centering
	\begin{minipage}{.44\textwidth}
		\centering
		\includegraphics[width=\linewidth]{img/head_specialization}
		\captionof{figure}{The attention heads specialized in different areas: left and front/right.}
		\label{fig:heads}
	\end{minipage}
\hfill
	\begin{minipage}{.545\textwidth}
		\centering
		\includegraphics[width=\linewidth]{img/distances}
		\captionof{figure}{The attention paid to a vehicle tends to increase as it gets closer.}
		\label{fig:distance}
	\end{minipage}
\end{figure}

@@

In this work, we showed that the \emph{list of features} representation, commonly used to describe vehicles in autonomous driving literature, is not tailored for use in a function approximation setting, in particular with neural networks. These concerns can be addressed by the \emph{spatial grid} representation, but it comes at the price of an increased input size and loss of accuracy. In contrast, we proposed an attention-based neural network architecture to tackle the aforementioned issues of the \emph{list of features} representation without compromising either size or accuracy. This architecture enjoys a better performance on a simulated negotiation and intersection crossing task, and is also more interpretable thanks to the visualisation of the attention matrix. The resulting policy successfully learns to recognize and exploit the interaction patterns that govern the nearby traffic.

>>>>

% ready for submission
% \usepackage[nonatbib]{neurips_2019_ml4ad}

@@

% to compile a camera-ready version, add the [final] option, e.g.:
\usepackage[final, nonatbib]{neurips_2019_ml4ad}

@@

\begin{abstract}
  We study the design of learning architectures for behavioural planning in a dense traffic setting. Such architectures should deal with a varying number of nearby vehicles, be invariant to the ordering chosen to describe them, while staying accurate and compact. We observe that the two most popular representations in the literature do not fit these criteria, and perform badly on an complex negotiation task. We propose an attention-based architecture that satisfies all these properties and explicitly accounts for the existing interactions between the traffic participants. We show that this architecture leads to significant performance gains, and is able to capture interactions patterns that can be visualised and qualitatively interpreted. Videos and code are available at \url{https://social-attention.github.io/}
\end{abstract}

@@

A popular way to address this limitations is to use a \emph{spatial grid} representation. Instead of explicitly representing spatial information as variables $x, y$ along with other features $f$ directly inside a state $\{s_i=(x_i,y_i,f_i)\}_{i\in[0,N]}$ indexed on the vehicles, they are instead represented implicitly through the layout of several feature variables $f_{ij}$ organised in a tensor structure, where the $(i,j)$ indexes refer to a quantisation of the 2D-space. This representation is illustrated in \autoref{fig:representation} (right). Note that the size of this tensor is related to the area covered divided by the quantisation step, which reflects a trade-off between accuracy and dimensionality.
In an occupancy grid, the $f$ features contains presence information (0-1) and additional channels such as velocity and heading, as in \citep[e.g.][]{Isele2018, Fridman2018, Bansal2018, Rehder2017c}. Another example is the use of top-view RGB images \citep[e.g.][]{Bagnell2010, Rehder2017, Rehder2017c, Liu2018}.

@@

The proposed architecture is presented in \autoref{fig:architecture}. It is used to represent the $Q$-function that will be optimized by the DQN algorithm. It is composed of a first linear encoding layer whose weights are shared between all vehicles. At that point, the embeddings only contain individual features of size $d_x$. They are then fed to an ego-attention layer, composed of several heads stacked together. The \emph{ego} prefix highlights that it similar to a multi-head self-attention layer \citep{Vaswani2017} but with only a single output corresponding to the ego-vehicle. Such an ego-attention head is illustrated in \autoref{fig:ego-attention} and works in the following way: in order to select a subset of vehicles depending on the context, the ego-vehicle  first emits a single query $Q = [q_0]\in\Real^{1 \times d_k}$, computed with a linear projection $L_q\in\Real^{d_x \times d_k}$ of its embedding. This query is then compared to a set of keys $K = [k_0, \dots, k_N]\in\Real^{N \times d_k}$ containing descriptive features $k_i$ for each vehicle, again computed with a shared linear projection $L_k\in\Real^{d_x \times d_k}$. The similarity between the query $q_0$ and any key $k_i$ is assessed by their dot product $q_0 k_i^T$. These similarities are then scaled by the inverse-square-root-dimension $1/\sqrt{d_k}$\footnote{This scaling is due to the fact that the dot-product of two independent random vectors with mean 0,  variance 1, and dimension $d_k$, is a random variable with mean 0 and variance $d_k$} and normalised with a softmax function $\sigma$ across vehicles. We obtain a stochastic matrix called the \emph{attention matrix}, which is finally used to gather a set of output value $V = [v_0, \dots, v_N]$, where each value $v_i$ is a feature computed with a shared linear projection $L_v\in\Real^{d_x \times d_v}$. Overall, the attention computation for each head can be written as:
\begin{equation}
\text{output}=\underbrace{\sigma\left(\frac{QK^T}{\sqrt{d_k}}\right)}_{\text{attention matrix}}V
\label{eq:selfattention}
\end{equation}
The outputs from all heads are finally combined with a linear layer, and the resulting tensor is then added to the ego encoding as in residual networks. We can easily see that this process is permutation invariant: indeed, a permutation $\tau$ will change the order of the rows in keys $K$ and values $V$ in \eqref{eq:selfattention} but will keep their correspondence. The final result is a dot product of values and key-similarities, which is independent of the ordering.

@@

In this application, we use the \href{https://github.com/eleurent/highway-env}{highway-env} environment \citep{highway-env} for simulated highway driving and behavioural decision-making. We propose a new task where vehicle-to-vehicle interaction plays a significant part: crossing a four-way intersection.
The scene -- composed of two roads crossing perpendicularly -- is populated with several traffic participants initialised with random positions, velocities, and destinations. As in \citep{highway-env}, these vehicle are simulated with the Kinematic Bicycle Model, their lateral control is achieved by a low-level steering controller tracking a target route, and their longitudinal behaviour follows the Intelligent Driver Model \citep{Treiber2000}. However, this model only considers same-lane interactions and special care was required to prevent lateral collisions at the intersection. To that end, we implemented the following simplistic behaviour: each vehicle predicts the future positions of its neighbours over a three-seconds horizon by using a constant velocity model. In case of predicted collision with a neighbour, the yielding vehicle is determined based on road priorities and brakes until the collision prediction ceases.

@@

We evaluate three different agents, whose characteristics are summarised in \autoref{tab:agents}.
\begin{itemize}
	\item \MLPL: a \emph{list of features} state representation is used, as described in \autoref{sec:background}. The model is a simple fully-connected network (FCN). Because this architecture requires a fixed-size input, we use zero-padding to fill the input tensor up to a maximum number $N=14$ of observed vehicles, and add an additional \emph{presence} feature to the coordinates described in \eqref{eq:coordinates} so as to identify active rows.
	\item \CNNG: a \emph{spatial grid} representation is used, as described in \autoref{sec:background}, with a $32 \times 32$ grid where each cell represents a $2\text{m}\times 2$m square. The model is a convolutional neural network (CNN).
	\item \EgoAtt: a \emph{list of features} state representation is used along with the Ego-Attention architecture described in \autoref{sec:architecture}. As this model supports varying-size inputs, zero-padding is not required.
\end{itemize}

\begin{table}[tp]
	\centering
	\begin{threeparttable}
		\caption{Characteristics of the agents}
		\label{tab:agents}
		\begin{tabular}{lccc}
			\toprule
			Architecture & \MLPL & \CNNG & \EgoAtt \\
			\midrule 
			Input sizes & [15, 7] & [11, 11, 7] & [1-15, 7] \\
			Layers sizes & [128, 128] &  \makecell[tc]{Convolutional layers: 3 \\ Kernel Size: 2 \\
			Stride: 2 \\ Head: [20]} & \makecell[tl]{Encoder: [64, 64] \\Attention: 2 heads, $d_k=32$ \\ Decoder: [64, 64]} \\
			Number of parameters & 3.0e4 & 3.2e4 & 3.4e4 \\
			Variable input size & No & No &  {Yes}  \\
			Permutation invariant & No & {Yes} &  {Yes} \\
			\bottomrule
		\end{tabular}
	\end{threeparttable}
\end{table}

@@

We plot in \autoref{fig:results} the evolution of the total reward, episode length and average velocity during training, over 4000 episodes and repeated across 120 random seeds.
The \MLPL agent learns to accelerate to earn short-term rewards, as shown by its high average velocity, but fails to exploit the information of other vehicles and crashes often, leading to short episodes. We obtain a risky and blind policy that is the worst performing.
Conversely, the \CNNG architecture benefits from its invariance to permutations and manages to learn to brake upon arrival at the intersection to avoid collisions, as we can see from its higher episode length. However, it only proceeds when the intersection has been fully cleared, as reflected by its low average velocity. This results in an overly cautious policy -- a common trait colloquially known as the \emph{freezing robot problem} \citep{Trautman2010} -- with a slight increase in performance.
In stark contrast, the \EgoAtt policy quickly learns both when it must slow down at the intersection (see the high episode length), but also when it can exploit the gaps in the traffic and take way to vehicles that are far or slow enough (see the higher average velocity than \CNNG). This translates as a significant performance improvement, and the overall resulting behaviour is qualitatively more nuanced and human-like.

@@

In any given state, the attention matrix can be visualised in the following way: we connect the ego-vehicle to every vehicle by a line of width proportional to the corresponding attention weight. Since the architecture can contain several ego-attention heads, we use different colours to distinguish them. In our experiments, two attention heads were used and will be represented in green and blue. We observe in \autoref{fig:heads} that they specialised to focus on different areas: the green head is only watching the vehicles coming from the left, while the blue head restricts itself to vehicles in the front and right directions. However, we notice that both heads exhibit a common behaviour: they direct their attention to incoming vehicles that are likely to collide with the ego-vehicle, depending on their current position, heading, velocity, and ignore those that are too far or in a conflict-less situation. In particular, the attention tends to increase when vehicles get closer, as shown in \autoref{fig:distance}. It can also be very sensitive to small variations in the traffic state, as reflected in \autoref{fig:sensitivity}. A full episode showcasing interactions with several vehicles is shown in \autoref{fig:episode}.

\begin{figure}[tp]
	\centering
	\begin{minipage}{.44\textwidth}
		\centering
		\includegraphics[width=\linewidth]{img/head_specialization}
		\captionof{figure}{The attention heads specialised in different areas: left and front/right.}
		\label{fig:heads}
	\end{minipage}
\hfill
	\begin{minipage}{.545\textwidth}
		\centering
		\includegraphics[width=\linewidth]{img/distances}
		\captionof{figure}{The attention paid to a vehicle tends to increase as it gets closer.}
		\label{fig:distance}
	\end{minipage}
\end{figure}

@@

In this work, we showed that the \emph{list of features} representation, commonly used to describe vehicles in autonomous driving literature, is not tailored for use in a function approximation setting, in particular with neural networks. These concerns can be addressed by the \emph{spatial grid} representation, but it comes at the price of an increased input size and loss of accuracy. In contrast, we proposed an attention-based neural network architecture to tackle the aforementioned issues of the \emph{list of features} representation without compromising either size or accuracy. This architecture enjoys a better performance on a simulated negotiation and intersection crossing task, and is also more interpretable thanks to the visualisation of the attention matrix. The resulting policy successfully learns to recognise and exploit the interaction patterns that govern the nearby traffic.
